---
title: "mckeithen-repro"
author: "Alissa McGill"
date: "`r Sys.Date()`"
output: pdf_document
---
---
title: "HLMFinalProject"
author: "Alissa McGill"
date: "2025-12-03"
output: pdf_document
geometry: margin=0.50in
options: width = 70
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Load Libraries, message=FALSE}

library(lme4)
library(lmerTest) #gives us some p values
library(haven) ## import ability to read sav (SPSS) files

```

```{r Load Data}

# mck_data =  read.csv("Myfile")

#create some dummy data for testing
set.seed(123)
mck_data <- data.frame(performance = rnorm(300),
                        codeVersion = factor(rep(numnbers[1:2], each = 150)),
                        experienceYears = double(rep(numbers[1:65], each = 100)),
                        trial = integer(rep(numbers[1,5]), each = 100)))


```


```{r Hsbdata}

#1)Create a model 
mck_model_1 = lmer(ses ~ 1 +  (1 | school), data = hsb_data, REML=FALSE )

summary(hsb_model_1)

#pull the variance values from thr output
lvl_1_variance = 0.4462
lvl_2_variance = 0.1597 

##ICC 
ICC = lvl_2_variance / ( lvl_1_variance + lvl_2_variance )
ICC

# Question:How much variability in socioeconomic status (the variable “SES” [not CSES] in the database) 
#is accounted for by the nesting of individuals within schools? What part of the output(s) did you use 
#to answer this question?

# Answer : The Interclass correlation (ICC) is the part of the output that tells us the variability that
#is accounted for by the nesting of individuals within schools. This ICC tells us that 26% of the 
#variability is explained by the grouping variable of "school".


#2)Create a model with only fixed effect for female
hsb_model_2 = lmer(ses ~ 1 + female +  (1 | school), data = hsb_data, REML=FALSE )

summary(hsb_model_2)

# Question: Does gender (“FEMALE”) predict socioeconomic status? Assuming that this variable
#is coded so that 0 = male, and 1 = female, what is the direction of this effect? What part of 
#the output(s) did you use to answer this question?

#Answer: Yes, gender ("FEMALE") predicts socioeconomic status - the p value for the "female"
#variable under fixed effects gives us a p-value of 6.37e-06 (0.000006). The estimate for the 
#coefficient is -0.083 , which means that for females (female=1), socioeconomic status is lower 
#then for non-females (female=0). 


#3a)Create a model with fixed effect and random effect for female
hsb_model_3 = lmer(ses ~ 1 + female +  (1 + female | school), data = hsb_data, REML=FALSE )

summary(hsb_model_3)

#Question: Does there appear to be systematic variability across schools in the effect of the gender?
#What part of the output(s) did you use to answer this question? Which model appears most defensible 
#given the observed data, the random effects model or the fixed effects only model?

#b3) deviance test to figure out if we care about the random effect: 
anova(hsb_model_1,hsb_model_2, hsb_model_3)

#because our Chi square is significant for model 2 - this tells us the random effect added in
#model 3 is NOT justified, 
#the fixed effect added in model 2 IS justified. 
#Choose model 2

#4) Test for moderation of school size

#create a new interaction term variable in the dataframe hsb
hsb_data$schSizeXfemale = hsb_data$size  * hsb_data$female

#create model with interaction term for school size and female (moderation)
hsb_model_4 = lmer(ses ~ 1 + female + size + schSizeXfemale + (1 | school), data = hsb_data, REML=FALSE )

summary(hsb_model_4)

# Question:  Does school size appear to moderate the effect of FEMALE on SES? What part of the output(s) 
#did you use to answer this question?

#The School size x female interaction term (schSizeXfemale ) was not significant (p=0.324), so school
#size does not moderate the effect of female on SES. 

#5) Test for moderation of Sector

#create a new interaction term variable in the dataframe hsb
hsb_data$sectorXfemale = hsb_data$sector  * hsb_data$female

#create model with interaction term for school size and female (moderation)
hsb_model_5 = lmer(ses ~ 1 + female + size + sectorXfemale + (1 | school), data = hsb_data, REML=FALSE )

summary(hsb_model_5)

#Question : Does school SECTOR (0 = public; 1 = private) appear to moderate the effect of FEMALE 
#on SES? What part of the output(s) did you use to answer this question?

#The sector x female interaction term (sectorXfemale ) was just barely not significant (p=0.0569 ), 
#so sector does not moderate the effect of female on SES. 

```

#Dataset 2 - HLM repeated
1)	How much variability in OUTCOME can be attributed to differences between individuals? 
What part of the output(s) did you use to answer this question?
2)	Does the variable AGE [do not transform or recode age for this step] predict differences
in OUTCOME? Assume that AGE is “years of age.” Please interpret the slope and the intercept.
3)	Does there appear to be systematic variability across persons in the effect of AGE? What
part of the output(s) did you use to answer this question? Which model appears most defensible
given the observed data, the random effects model or the fixed effects only model?
4)	Now, using the model that you chose in part 3, test the following: Does FEMALE appear to 
moderate the effect of AGE on OUTCOME? What part of the output(s) did you use to answer this question?
```{r HLM Repeated dataset}

#1)Create a model with just the nesting of the individual
rep_model_1 = lmer(outcome ~ 1 +  (1 | id), data = repeated_data, REML=FALSE )

summary(rep_model_1)

lvl_1_variance = 4.930
lvl_2_variance = 3.567

##ICC 
ICC = lvl_2_variance / ( lvl_1_variance + lvl_2_variance )
ICC

# Question: How much variability in OUTCOME can be attributed to differences between 
#individuals? What part of the output(s) did you use to answer this question?

# 42% of the variance in outcome is explained by the individual. I calculated and used 
#the ICC to determine this.

#2)Create a model with only fixed effect for age
rep_model_2 = lmer(outcome ~ 1 + age +  (1 | id), data = repeated_data, REML=FALSE )

summary(rep_model_2)

#Question : Does the variable AGE predict differences in OUTCOME?
#Answer: Yes, Age predicts differences in OUTCOME. We can see in the fixed effects
#table that p=<2e-16. ... which is significant 

#Question:Please interpret the slope and the intercept.
#Answer: The intercept is the value of outcome when the variable age=0. The
#slope means that for every 1 point increase in age, the OUTCOME increases by 0.66.  


#3a)Create a model with fixed effect and random effect for age
rep_model_3 = lmer(outcome ~ 1 + age +  (1 + age | id), data = repeated_data, REML=FALSE )

summary(rep_model_3)

#b3) deviance test
anova(rep_model_1,rep_model_2, rep_model_3)

#Does there appear to be systematic variability across persons in the effect of AGE? What part
#of the output(s) did you use to answer this question? #Which model appears most defensible given
#the observed data, the random effects model or the fixed effects only model?

# Similar to our previous question, because our Chi square is significant for model 2 (p=<2e-16) - 
#this tells us the random effect added in model 3 is NOT justified, 
#but the fixed effect added in model 2 IS justified. 
#Choose model 2 again!


#4) Test for moderation of Female

#create a new interaction term variable in the dataframe repeated
repeated_data$femaleXage = repeated_data$female  * repeated_data$age

#create model with interaction term for school size and female (moderation)
rep_model_4 = lmer(outcome ~ 1 + age + female + femaleXage + (1 | id), data = repeated_data, REML=FALSE )

summary(rep_model_4)

# Yes, we do see that female moderates in effect of age on OUTCOME. The interaction term 
#(femaleXage) had a p value of p=0.0129, which is significant, showing that we do have moderation.    

